\label{sec:intro}

\textbf{Position paper.}
This non-archival submission responds to the Rethinking Safety workshop call
for forward-thinking perspectives---not a complete systems paper.
Our \textbf{core claim} is:
\begin{quote}
\small
\emph{Generalist-robot safety must include initiation authorization: deciding
whether a robot may commit its first irreversible social act under uncertainty.}
\end{quote}
\noindent
Initiation is not engagement; it is authorization under partial observability
\citep{rich2010,salam2017,broek2024proactive,shi2011spatial}.
We are \emph{not} proposing a new collision checker, VLA policy, or dialogue
model.

\paragraph{The conceptual gap.}
Three safety layers are often conflated:
\textbf{physical safety} constrains motion \citep{zhang2025safevla,kojima2025fmr};
\textbf{social-interaction safety} governs quality and norms \emph{during}
contact \citep{akalin2021,weiss2023perceived,tian2021socialerrors};
\textbf{initiation authorization}---largely missing from reference
architectures---governs whether to open the channel at all.
The workshop's context-dependent knife/egg examples show that ``safe'' depends
on context and preference; the same structure applies beyond reception robots:
a spoken greeting can interrupt a phone call \citep{edirisinghe2024shopworker};
a home generalist can speak over a human conversation; a VLA can reach for a
contested object before assent \citep{rt2-2023,openvla-2024,black2024pi0};
jailbroken planners can authorize harmful \emph{first moves} as readily as
harmful first words \citep{robey2024jailbreaking,ravichandran2025roboguard}.

Most deployed stacks nevertheless implement initiation as a fixed threshold,
$\texttt{ACT/SPEAK} \Leftrightarrow e_t > \theta$ \citep{bohus2014}---treating
engagement score or a single VLA rollout as social permission without context,
preference, or bidirectional evidence from low-cost probes
\citep{skantze2014speech,vasquez2014,bohus2014}.
Figure~\ref{fig:teaser} contrasts this pattern with initiation safety as
\emph{safety-by-practice}: reversible probes before irreversible commitment,
explicit preference, and runtime shields before foundation-model dialogue.

\input{text/figure1}

\paragraph{Contributions.}
\begin{enumerate}\setlength{\itemsep}{1pt}
  \item \textbf{Taxonomy:} position initiation authorization as a critical
  subclass of social-interaction safety---distinct from physical shields, cyber
  risk, post-dialogue alignment \citep{bai2022constitutional}, and engagement
  optimization.
  \item \textbf{Framing:} relate workshop themes to pre-commitment decisions
  under partial observability, latent receptivity, and asymmetric social costs.
  \item \textbf{Discussion catalyst:} sketch PAS (probe--authorize--speak) on a
  real robot, state an evaluation protocol, and pose open questions for
  community standards (quantitative outcomes \textbf{in progress}, $n{=}90$).
\end{enumerate}
\noindent
Section~\ref{sec:taxonomy} develops the taxonomy; Section~\ref{sec:pas} gives an
illustrative sketch and protocol; Section~\ref{sec:open} poses open questions and
provisional statements; Section~\ref{sec:conclusion} closes.
