\subsection{Future Work}
\label{sec:futurework}

Our immediate next step is to complete the between-subjects study
($n{=}90$) and analyze whether margin distributions differ across
\textsc{pas}, direct-init, and passive-wait, and whether margin
alignment tracks user-reported awkwardness. Beyond that, three
directions follow naturally. First, we plan to extend \textsc{pas} from
single-encounter doorways to multi-party scenes, where addressee
selection becomes part of authorization. Second, we plan to integrate
\textsc{pas} with VLA action policies, so that initiation authorization
gates not only speech but reach and approach. Third, we aim to release
the initiation log format as an open schema, supporting cross-platform
comparison of first-act decisions.

\subsection{Open Questions for the Community}

Retail, home, and museum settings differ but share one pattern: the
robot acts before knowing who is willing to be addressed
\citep{edirisinghe2024shopworker,shi2011spatial}. This raises three
families of open questions.

\paragraph{Measurement and standards.} Should ``acting too soon'' appear
explicitly in standard safety designs, or stay buried inside
engagement metrics? What minimum metrics (awkwardness, interruptions,
correct addressee, logged margins) let us compare robots across
venues? Without metrics for the first act, a platform can look safe on
average while still often opening with a harmful first word in the
scenes where initiation matters \citep{akalin2021}.

\paragraph{Preference and governance.} Who sets how bold or conservative
the robot should be: operator, user, online system, or regulator? Can
one rule set cover all generalist robots, or do we need safety cases
per domain? A home user may want proactive help, while a clinic may
need conservative initiation. Generalist stacks need a setting that
operators can understand and audit after incidents, not a single
threshold frozen in model weights.

\paragraph{Foundation models and layering.} If dialogue models improve
fluency but not timing, where should generative planning end and
runtime checks begin \citep{ravichandran2025roboguard}? Which failures
(wrong addressee, jailbroken greeting, speaking too much too soon)
appear only when initiation and generation are merged? Red-teaming
should cover the first physical act, where jailbreaks appear as harmful
first moves rather than toxic mid-dialogue replies---problems that
text-only tests miss \citep{robey2024jailbreaking}. End-to-end VLAs
\citep{openvla-2024} make this urgent because one rollout can tie
together perception, commitment, and action.

\subsection{Provisional Statements}

\paragraph{Statement 1: Initiation is safety-critical.} For social
generalists, deciding when to start interaction is a safety decision
about hard-to-undo actions---similar in structure, if not in physics,
to the familiar egg/knife context examples.

\paragraph{Statement 2: Hard-to-undo acts need explicit
authorization.} Starting speech, naming a user, or beginning
biometric capture should not flow unchecked from a single engagement
score or an end-to-end VLA. Runtime gates, probes, and auditable
margins belong in standard designs alongside motion shields.

\paragraph{Statement 3: Initiation logs should be required at
deploy time.} Generalist-robot safety standards should require logged
initiation decisions (probe history, authorization margins, time of
first act), similar to incident logs in aviation. Without auditable
records of the first act, ``felt unsafe'' reports cannot improve policy
after a harmful first word or first grasp.
