We implement initiation authorization as \textsc{pas}
(probe--authorize--speak) on SocialVLA-Engage. \textsc{pas} deliberately
uses familiar HRI primitives---an engagement score, a probe FSM, and a
threshold gate \citep{bohus2014,skantze2014speech,rich2010}, arranged as
a safety layer with (i) a probe module structurally prevented from
emitting \textsc{speak}, (ii) a gate threshold $\tau(\rho)$ that is a
deploy-time dial rather than a fixed trained constant, and (iii) a
logged margin $\Delta_{\text{init}}$ at every first-\textsc{speak}
event. The controller runs at 10\,Hz. Only the authorization gate may
release first speech, and dialogue generation runs after that point.

\paragraph{Runtime split.} Each cycle has three steps.
\textbf{(1) Perception} tracks visible people and fuses gaze, speech
activity, approach, and dwell into an engagement score $e_t$.
\textbf{(2) Probe FSM} moves through four stages of nonverbal signaling.
At each stage the robot may only \emph{wait}, \emph{turn its head toward
the person}, or \emph{turn its body to face the person}---never speak.
A higher stage requires stronger signs that the person is open to
interaction; if the person looks away, the stage can drop back.
In Table~\ref{tab:probe_stages}, the middle column lists what the
person must show to \emph{enter} that stage; the right column is what
the robot does there. Speech is decided in step~(3).
\begin{center}
\refstepcounter{table}\label{tab:probe_stages}
\footnotesize
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}c>{\raggedright\arraybackslash}p{0.36\columnwidth}>{\raggedright\arraybackslash}p{0.44\columnwidth}}
\toprule
\textbf{St.} & \textbf{To enter stage} & \textbf{Robot does} \\
\midrule
0 & Someone nearby & Waits; stays still \\
1 & Brief eye contact & Turns head toward person \\
2 & Sustained attention; not walking away & Turns body to face person \\
3 & Lingers nearby & Keeps facing person; gate may allow speech \\
\bottomrule
\end{tabular*}
\end{center}
\textbf{(3) Authorization gate} opens first speech only when
$e_t \geq \tau(\rho)$ and probe stage $\geq 2$, where $\tau(\rho)$ is
set by the venue bold/conservative dial $\rho$
\citep{harland2024adaptivealignment}. The system logs the
margin $\Delta_{\text{init}} = e_t - \tau(\rho)$ at first
\textsc{speak}.

\begin{algorithm}[h]
\caption{PAS initiation loop (10\,Hz)}
\small
\begin{algorithmic}[1]
\State \textbf{Input:} frame, stage, $\rho$
\State $e_t \leftarrow \text{engagement\_score}()$
\State $\tau \leftarrow \tau(\rho)$
\State $a_t \leftarrow \text{probe\_fsm}(e_t,\text{stage})$
\Comment{never \textsc{speak}}
\If{$e_t \geq \tau$ \textbf{and} stage $\geq 2$}
  \State $a_t \leftarrow$ \textsc{speak}
  \State log $\Delta_{\text{init}} = e_t - \tau$
  \State run FM dialogue
\EndIf
\end{algorithmic}
\end{algorithm}

\paragraph{What is different from engagement-based initiation.}
Prior engagement-based initiators emit speech once the score crosses a
tuned threshold \citep{bohus2014,rich2010}. \textsc{pas} makes three
changes that prior work does not separate or log. First, the probe
module is structurally prevented from emitting \textsc{speak}---the
safety property is enforced in the controller, not left to a learned
policy. Second, $\tau(\rho)$ is a deploy-time dial: the same perception
stack can support a conservative museum setting and a bolder retail demo
without retraining. Third, the system logs $\Delta_{\text{init}}$ at
every first-\textsc{speak} event, making initiation decisions auditable
after incidents.

\paragraph{What the study changes.} All conditions share the same
humanoid, perception stack, doorway scenario, and post-initiation
dialogue. Only the initiation layer differs.
\textsc{pas} uses the full probe--gate split with default $\rho$.
Direct-init maps $e_t > \theta$ directly to \textsc{speak}, skipping
probes and the gate.
Passive-wait keeps probes but uses a conservative $\rho$ so the gate
rarely opens.

\paragraph{Study scenario.} Participants meet the robot at a scripted
doorway. Each trial scripts one of three visitor states: on a phone,
mid-conversation, or clearly open to interaction. This study
instantiates the framing for doorway speech. Reach, approach, and
multi-party authorization are deferred to future work (\S\ref{sec:futurework}).
Sessions log probe actions, gate state, stage, $\Delta_{\text{init}}$ at
first \textsc{speak}, and user-reported awkwardness, success, and
naturalness.

\subsection{Evaluation Protocol}

We compare these three policies in a between-subjects study
($n{=}90$, ethics-approved):

\begin{center}
\small
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{0.28\columnwidth}>{\raggedright\arraybackslash}p{0.66\columnwidth}@{}}
\toprule
\textbf{Condition} & \textbf{What it does} \\
\midrule
\textsc{pas} & Probes + gate (default $\rho$) \\
\midrule
Direct-init & $e_t > \theta$, no probes or gate \\
\midrule
Passive-wait & Probes + gate with conservative $\rho$ \\
\bottomrule
\end{tabular*}
\end{center}

\paragraph{Endpoints.} \emph{System-level:} distribution of
$\Delta_{\text{init}}$ at first \textsc{speak}, comparing \textsc{pas}
with direct-init. \emph{User-reported:} conversation success,
awkwardness, and naturalness on 7-point Likert scales.
\emph{Deployment:} sweep $\rho$ on \textsc{pas} without retraining
perception.

\paragraph{Ways the framing could fail.} The claim is testable: if
\textsc{pas} and direct-init produce the same $\Delta_{\text{init}}$
distributions, the probe--gate split adds little.
If awkwardness does not track margin alignment, felt safety likely
needs metrics beyond timing proxies.
If sweeping $\rho$ does not trade speaking rate against margin in a
clear way, tunable authorization is not a useful deploy-time dial.
The study tests whether initiation authorization can be tuned and
whether margins differ across policies.
