
\subsection{Where Initiation Fits}
\label{sec:placement}

Physical safety---collisions, forces, joint limits---remains
necessary
\citep{akalin2021,zhang2025safevla}, and cyber and
privacy safety---data leakage, jailbreak prompting---also matter for
foundation-model robots
\citep{robey2024jailbreaking,bai2022constitutional}. Beyond these,
robots that initiate contact in public or home settings face a third
question: should they act at all?

HRI already separates physical safety from how safe people feel
\citep{akalin2021}, and social-error taxonomies
list what goes wrong once interaction is already underway
\citep{tian2021socialerrors}. What is missing is a layer for the
earlier decision of whether the robot should start in the first place
\citep{shi2011spatial,edirisinghe2024shopworker}. We call this layer
\emph{initiation authorization} and treat it as a key part of
social-interaction safety.

\begin{table}[H]
\footnotesize
\centering
\caption{Safety layers for generalist robots.}
\label{tab:taxonomy}
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{0.21\columnwidth}>{\raggedright\arraybackslash}p{0.37\columnwidth}>{\raggedright\arraybackslash}p{0.26\columnwidth}@{}}
\toprule
\textbf{Layer} & \textbf{What it protects} & \textbf{Example problem} \\
\midrule
Physical & Contact force, clearance & Unintended touch \\
\midrule
Cyber / privacy & Access control, private data & Leaked session data \\
\midrule
Social-interaction & Felt safety, social norms & Interrupt, pressure \\
\midrule
\quad$\hookrightarrow$ Initiation & When to act, whom to address & Acting too soon \\
\bottomrule
\end{tabular*}
\end{table}

Initiation is not only about speech. On talking robots the first
hard-to-undo action is often a word. On manipulation VLAs it can be
the first grasp, stepping into someone's space, or a question that
cuts into a human conversation \citep{openvla-2024}.
What matters is whether the action is hard to take back while the
situation is still unclear---not whether it is speech, reach, or
approach. Table~\ref{tab:themes} maps workshop safety concerns to
these before-the-first-act decisions.

\begin{table}[H]
\small
\centering
\caption{Workshop safety themes, mapped to initiation.}
\label{tab:themes}
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{0.34\columnwidth}>{\raggedright\arraybackslash}p{0.60\columnwidth}@{}}
\toprule
\textbf{Concern} & \textbf{Question before the first act} \\
\midrule
Venue and context & Same greeting or reach OK in demo, rude at home \\
\midrule
Willingness to interact & Looking at the robot $\neq$ wanting to be greeted \\
\midrule
Social risk preference & Bold greeting vs.\ visitor who wants privacy \\
\midrule
Speaking too early vs.\ too late & Awkward interrupt vs.\ missed visitor \\
\midrule
Speak, act, and log permissions & Who may speak, move, log, or learn from data \\
\bottomrule
\end{tabular*}
\end{table}

\FloatBarrier

\subsection{From Specialist to Generalist Safety}

Single-purpose robots are often built on assumptions that generalists
cannot use: (i) one threshold $\theta$ works for everyone,
(ii) an engagement score is enough to know intent, with no probe,
(iii) speaking too soon and waiting too long are equally bad mistakes,
and (iv) the same script works in every venue. Generalists need
\emph{tunable authorization}: a per-venue and per-user setting for
how bold or conservative to be, without retraining the whole VLA stack. Table~\ref{tab:specialist} contrasts the two
settings.

\begin{table}[H]
\small
\centering
\caption{How specialist and generalist robots differ on safety.}
\label{tab:specialist}
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{0.24\columnwidth}>{\raggedright\arraybackslash}p{0.35\columnwidth}>{\raggedright\arraybackslash}p{0.35\columnwidth}@{}}
\toprule
\textbf{Question} & \textbf{Specialist robot} & \textbf{Generalist robot} \\
\midrule
What goes wrong? & Collisions, excess force & Broken norms, bad timing \\
\midrule
What can the robot see? & Positions and clearance & Whether the person wants contact---not directly readable \\
\midrule
What counts as safe? & Stay inside motion limits & Sometimes do nothing at all \\
\midrule
How is it tuned? & Calibrate once at deploy & Adjust per venue or user \\
\bottomrule
\end{tabular*}
\end{table}

\subsection{Layered Safety and Adjacent Literatures}

If specialist assumptions fail, one safety layer is not enough. A
full stack needs three checks. Physical shields and motion planning
limit contact risk. VLA alignment
\citep{zhang2025safevla} checks whether a chosen plan
is safe once the robot is already trying to act. \emph{Initiation
authorization} \citep{ravichandran2025roboguard} runs before the first
hard-to-undo social action. Recent post-plan guardrails catch unsafe or
jailbroken plans, but only at that later stage. They do not decide
whether the robot should speak, reach, or approach in the first place
\citep{ravichandran2025roboguard,robey2024jailbreaking}.
We now separate initiation authorization from three nearby lines of
work.

\paragraph{vs.\ physical safety and VLA guardrails.} Collision
avoidance and SafeVLA-style guardrails ask: if the robot acts, will
this motion hurt someone? \citep{zhang2025safevla} Initiation
authorization asks an earlier question: should the robot act now at
all? A collision-free robot can still be socially harmful. A near-miss
with a knife is the physical analogue
\citep{akalin2021}.

\paragraph{vs.\ preference alignment and LLM safety.} Preference
alignment shapes what a model says or does after interaction has
started \citep{bai2022constitutional}. Before the first act is
different: small probes can change how the person responds.
Harm is social and goes both ways, and the hard-to-undo action comes
before the person can correct it. Red-teaming a dialogue model does not
replace testing probe--response behavior on a physical robot
\citep{robey2024jailbreaking}.

\paragraph{vs.\ engagement and proactive HRI.} Proactive-robot
research studies \emph{how} to initiate well---timing, position,
naturalness
\citep{broek2024proactive,shi2011spatial}.
Engagement metrics maximize interaction quality
\citep{rich2010}. We study \emph{whether} initiating is
allowed when the situation is still unclear. Engagement scores are not
consent. A safe policy must allow the robot to wait when evidence is
weak. The mechanism-level consequence is that probes must be incapable
of triggering speech, the gate must be tunable per venue, and margins
must be logged---properties outside what engagement metrics or
proactive-HRI timing models are designed to provide. A policy tuned
for high engagement may act early. A safe initiation policy may not.
